Functional trait data tell new story about biodiversity change

### 2021-04-15 paper/manuscript.Rmd
Clean up methods, add discussion

Ecological communities are experiencing unprecedented change as a result
of anthropogenic pressures such as climate change, land use change, and
invasive species. Impacts of these pressures are well documented at a
global scale by an accelerating global extinction rate [@barnosky2011],
and fundamental changes in some of the well-studied systems (e.g. coral
bleaching). At the local scale however, species diversity tells a
different story. Recent synthesis of local trends in biodiversity over
time have found no net change in local biodiversity despite ongoing
species turnover [@brown2001; @dornelas2014; @vellend2013; @vellend2017]
and evidence of a decoupling of community composition from richness
[@brose2016; [@gotelli2017]. While communities are clearly changing, our
most common species-based approaches do not fully capture the nature of
that change.

Functional diversity offers a potentially powerful alternative for
detecting and describing community change by providing a mechanistic
link between the organisms in a system and the processes they perform
[@mcgill2006]. By summarizing functional traits rather than species,
functional diversity metrics capture the disproportionate impact of
losses or gains of functionally unique species. Functional diversity
metrics are particularly suited for understanding community-level
biodiversity patterns as they can detect shifts in community structure
independent of species change.

Functional diversity sheds further light on critical ecosystem and
community processes potentially impacted by anthropogenic change. For
example, there is increasing evidence that functional diversity is a
better predictor of ecosystem function than species-based metrics
[@cadotte2011; @gagic2015], and that different facets of functional
diversity also play a role in maintaining ecosystem stability
[@craven2018; @morin2014]. Further, biogeographic assessments of
functional diversity give foundational insight into how community
assembly plays out across systems at a standardized scale [@violle2014;
@weiher2011].

@@

Assessments of broad-scale temporal change in functional diversity have
previously been limited by a lack of functional trait data. The majority
of work has therefore focused largely on system-specific studies with
traits collected *in situ*. Ongoing efforts to assemble functional
traits for a variety of taxa have made synthesis of existing community
assemblage data and functional traits possible for the first time,
providing initial looks at how functional diversity changes on a
broad-scale for specific taxa [@trindade-santos2020; @jarzyna2016;
@barnagaud2017].
<!-- so what do these studies say? does that go here? --> However, to
date there has been no cross-taxa assessment of functional change
through time on a broad geographic and taxonomic extent.\

Here we perform the first multitaxa, multisystem assessment of
functional diversity change through time. We assess 100's of mammal,
bird, and amphibian functional diversity timeseries to 1) determine how
functional diversity change compares to species change and 2) identify 3
distinct types of functional change common across communities.

### Methods

@@

We obtained mammal, bird, and amphibian timeseries from the Biotime
database, a global repository of high quality community assemblage
timeseries. All studies included in the database follow consistent
sampling protocols and represent full assemblages rather than
population-level data [@dornelas2018]. Following best practices for the
database [@blowes2019] studies with multiple sample locations were split
into individual timeseries following a standardized scale. A global grid
was defined with a cell size determined based on the sample extent of
studies with only a single location (see [@dornelas2018] for details on
how sample extents were defined), with the area of each cell set to one
standard deviation away from the mean of the single extent locations.
All samples from a study within a single cell were considered to be a
single timeseries, and species abundances were combined for all samples.

We gathered trait data from the Elton Trait Database (mammals and birds;
[@wilman2014]) and Amphibio (amphibians, [@oliveira2017]). These
databases include species-level means for traits that best define
species' function in the community including body size, diet, and
behavioral characteristics <!-- what's in the amphibian database? -->.
For the full list of traits included in the analysis see the
supplemental materials.

@@

#### *Functional diversity metrics*

We calculated yearly metrics of functional diversity and turnover for
each timeseries within each study. All available trait data for each
study was included in the trait matrix with the exception of traits that
were the same value for all observed species in the study. All
continuous traits were z-score scaled to give each trait equal weight in
the trait space [@leps2006; @schleuter2010]. We also calculated basic
species richness, *S.*

We used the function dbFD from the FD package (citation) to calculated
functional diversity (FD), functional evenness (FEve), functional
diveregence (FDiv), and functional dispersion (FDis). The number of
trait axes was limited to the maximum number of traits that fulfills the
criteria $s >= s^t$, where $s$ is the number of species and \$t\$ is the
number of traits. This restriction allows for a sufficient number of
axes to capture the trait space while maintaining computational
feasibility for metric calculations. Metrics incorporated weighting
based on either species abundance or biomass based on the methods for
each study.

#### ***Null Models***

In order to control for changes in functional diversity metrics due to
species richness, we calculated the standardized effect size (SES) for
each metric from null estimates [@swenson2012]. For each richness level
observed in a timeseries, our null model randomly selected the same
number of species from the species pool, which included all species
observed over the course of the timeseries. We then randomly assigned
abundances from the original sample to each species, and all functional
diversity metrics were calculated. This process was repeated 500 times
to get a estimate and standard deviation of the null expectation for the
metric for that timeseries at that richness level. We used these values
to calculated SES using the following formula:
$SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}$.

@@

We modeled each diversity metric using a Linear Mixed Effects model with
a random slope and intercept for year and a nested random effect of
timeseries nested within study.

>>>>

Ecological communities are experiencing unprecedented change as a result
of anthropogenic pressures such as climate change, land use change, and
invasive species. Impacts of these pressures are well documented at a
global scale by an accelerating global extinction rate [@barnosky2011],
and fundamental changes in some of the most well-studied systems (e.g.
coral bleaching). At the local scale however, species diversity tells a
different story. Recent synthesis of local trends in biodiversity over
time have found no net change in local biodiversity despite ongoing
species turnover [@brown2001; @dornelas2014; @vellend2013; @vellend2017]
and evidence of a decoupling of community composition from richness
[[@brose2016]; [@gotelli2017]. Potential hypothesis for the balance of
etc..<!-- potential indication of biological homogenization due to shifting ranges and/or invasive species-->
While communities are clearly changing, our most common species-based
approaches do not fully capture the nature of that change.

Functional diversity offers a potentially powerful alternative for
detecting and describing community change by providing a mechanistic
link between the organisms in a system and the processes they perform
[@mcgill2006]. By summarizing functional traits rather than species,
functional diversity metrics capture the disproportionate impact of
losses or gains of functionally unique species. Functional diversity
metrics are therefore particularly well suited for addressing change in
communities with constant species richness.

Functional diversity also has the potential to shed light on critical
ecosystem and community processes potentially impacted by anthropogenic
change. For example, there is increasing evidence that functional
diversity is a better predictor of ecosystem function than species-based
metrics [@cadotte2011; @gagic2015], and that different facets of
functional diversity also play a role in maintaining ecosystem stability
[@craven2018; @morin2014]. Further, biogeographic assessments of
functional diversity may give foundational insight into how community
assembly plays out across systems [@violle2014; @weiher2011].

@@

<!-- paragraph about what we already know about global functional changes -->

Assessments of broad-scale temporal change in functional diversity have
previously been limited by a lack of functional trait data. The majority
of work has therefore focused largely on system-specific studies with
traits collected *in situ*. Ongoing efforts to assemble functional
traits for a variety of taxa have made synthesis of existing community
assemblage data and functional traits possible for the first time,
providing initial looks at how functional diversity changes on a
broad-scale for specific taxa (e.g. fish[@trindade-santos2020], birds
[@jarzyna2016; @barnagaud2017]).
<!-- so what do these studies say? does that go here? --> However, to
date there has been no cross-taxa assessment of functional change
through time across a broad geographic and taxonomic extent.

Here we perform the first multitaxa, multisystem assessment of
functional diversity change through time. We assess thousands of mammal,
bird, and amphibian functional diversity timeseries to 1) determine how
functional diversity change compares to species change and 2) identify
three distinct types of functional change common across communities.
While our data represent the largest collection of functional diversity
timeseries to date, they are not a representative sample of systems
experiencing biodiversity change. We do not therefore seek to classify
the global trend in functional diversity change, but rather assess the
utility of a functional lens in understanding biodiversity change.

### Material and Methods

@@

We obtained mammal, bird, and amphibian timeseries from the Biotime
database, a global repository of high quality community assemblage
timeseries. All studies included in the database follow consistent
sampling protocols and represent full assemblages rather than population
of single species. [@dornelas2018]. Following best practices for the
database [@blowes2019] studies with multiple sample locations were split
into individual timeseries following a standardized scale. A global grid
was defined with a cell size determined based on the sample extent of
studies with only a single location (see [@dornelas2018] for details on
how sample extents were defined), with the area of each cell set to one
standard deviation away from the mean of the single extent locations.
All samples from a study within a single cell were considered to be a
single timeseries, and species abundances were combined for all samples.

We gathered trait data from the Elton Trait Database (mammals and birds;
[@wilman2014]) and Amphibio (amphibians, [@oliveira2017]). These
databases include species-level means for traits that best define
species' function in the community including body size, diet, and
behavioral characteristics <!-- what's in the amphibian database? -->.
For the full list of traits included in the analysis see the
supplemental materials.
<!-- limitations of species level trait data? -->

@@

Our final dataset included 2439 timeseries from 54 studies in 21
countries and 15 biomes. Only four studies (consisting of 11 timeseries)
come from Amphibian studies due to the limited availability of amphibian
timeseries and low species richness values in those timeseries. For a
full breakdown of studies and their characteristics see the supplement.

#### *Diversity Metrics*

We calculated yearly metrics of functional and species diversity for
each timeseries. Species-based metrics include species richness (*S*)
and Jaccard similarity (*J*) as a measure of turnover. Jaccard
similarity was calculated relative to the first observed year for a
timeseries. A negative trend would therefore indicate increasing
turnover.

Functional diversity metrics were calculated using the dbFD function
from the FD R package. Here we report functional richness (*FRic*),
functional evenness (FEve), and functional diveregence (FDiv) which
together describe three complementary characteristics of the functional
space [@mason2005]. FRic assesses the volume of the trait space occupied
by species in the community. FEve describes how species are distributed
across the trait space and how abundance is distributed across species.
FDiv measures the degree to which species and their abundances maximize
differences in the functional space. Higher values of FDiv therefore
correspond to communities where many highly abundant species are on the
edges of the trait space.

All available trait data for each study were included in functional
diversity calculations with the exception of traits that were the same
value for all observed species in the study. All continuous traits were
z-score scaled to give each trait equal weight in the trait space
[@leps2006; @schleuter2010]. The number of trait axes was limited to the
maximum number of traits that fulfills the criteria $s >= s^t$, where
$s$ is the number of species and \$t\$ is the number of traits. This
restriction allows for a sufficient number of axes to capture the trait
space while maintaining computational feasibility. Metrics incorporated
weighting based on species abundance where available.

#### *Null Models*

We corrected for bias in functional diversity metrics due to changes in
species richness by calculating the standardized effect size (SES) for
each metric from null estimates [@swenson2012]. Null model corrections
allow us to assess the degree to which the observed functional diversity
metric deviates from the value expected by chance in a randomly
assembled community. Null estimates were calculated for each rarefied
sample by randomly sampling species from the species pool for each year
and randomly assigning observed abundances to species. Species pools
included all species observed for a timeseries. This process was
repeated 500 times to get an estimate and standard deviation of the null
expectation for the metric for that timeseries. We used these values to
calculate SES using the following formula:
$SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}$. Unless otherwise
specified, functional diversity results are reported for standardized
values.

@@

We estimated overall trends for each diversity metric using a Linear
Mixed Effects model with a random slope and intercept for each
timeseries nested within study. Differences in trends by taxa, biome,
and realm were also modeled individually as covariates in models of the
same structure.

We assessed the type of changes happening in studies across metrics by
performing a cluster analysis of the study-level trends for each metric
using hierarchical clustering. Studies were clustered based on the
overall temporal slope for each metric. We included all standardized
functional diversity metrics and species richness to represent
independent facets of biodiversity change.
<!-- why do some studies not have blup estimates for FRic?-->

### Results

We found no significant temporal trend in species richness or functional
diversity metrics for either observed or standardized metrics. The lack
of trend also held when data was broken down by taxa, biome and realm.
We did find a significant decrease in Jaccard similarity, indicating
increasing turnover through time.

At the study level, the majority of studies (78%) experienced a
significant temporal change in species richness and all functional
diversity metrics. Only four studies experienced no significant trend in
species richness clustering analysis showed that all but a few sites
were part of a cluster characterized by minimal deviance from a zero
temporal trend estimate for all metrics (Table 1). We identified four
other clusters containing six studies that were characterized by outlier
trends in at least one of the metrics and near-zero trends in the rest.
The first contained two studies with the most strongly decreasing
evenness. The second, two studies with the most strongly increasing
functional divergence and increasing functional richness. The third
contained one study with strongly decreasing functional divergence and
richness, and the fourth strongly decreasing species richness with
increasing functional richness.

### Discussion

Our study represents the first broad-scale multi-taxa assessment of
functional change through time, giving a first look at overall trends in
functional diversity and the range of changes happening in mammal, bird,
and amphibian communities.

We found no systematic temporal change in any measured dimension of
functional diversity or species richness, even within specific realms,
biomes, or taxonomic groups. Species-based results are consistent with
other broad scale analyses of biodiversity change through time
[@dornelas2014] that found no overall trend in species richness despite
increasing turnover. While nearly all of the studies included in our
data experienced a significant increase or decrease in species richness,
they were observed in equal numbers resulting in an overall trend not
significantly different from zero. Surprisingly, functional diversity
metrics also did not capture any overall trends, indicating that similar
to species richness, communities are experience a wide range of
functional shifts that cannot be explained by their taxa, realm, or
climate.

While species-level metrics are the only functional data available to
perform an analysis of this scale, they are unable to detect functional
shifts in communities with little to no turnover as they do not capture
intraspecific shifts.

Data limitations:

-   species-level traits

-   not a lot of changes in these systems already?

#### Add to supplement

-   breakdown of number of studies/timeseries for each taxa/realm etc
